---
title: "What makes a successful Formula One team?"
author: by ECK
output: github_document
---

```{r, libraries-and-data, echo = FALSE, message = FALSE}
library(tidyverse)
library(broom)
library(here)
library(skimr)
library(glue)
library(forcats)

f1merged <- read_csv("/cloud/project/data/f1merged.csv")
f1merged_hybrid <- read_csv("/cloud/project/data/f1merged_hybrid.csv")
```


## Summary

Our project is to investigate what makes a good Formula One driver? Specifically, we are looking at the Mercedes-AMG Petronas team in the hybrid era (2014-2020) and searching for possible reasons for their success.

To answer the question "What makes a successful Formula One team?", we want to see what variables in the data could impact a drivers performance. Since our data is actually a combination of multiple datasets, we combined the datasets *results, races, drivers* and *constructors* and focused only on entries in the hybrid era (excluding 2021 since the 2021 races are ongoing at the time of this analysis). This resulting dataset was named *f1merged_hybrid*.

Firstly, we wanted to investigate the performance of the Mercedes team. We looked at the total number of wins of five key teams - AlphaTauri, Ferrari, Mercedes, Racing Point and Red Bull. 

```{r, wins-bar-chart, echo = FALSE}
team_colours <- c("Mercedes" = "#00d2be",
                  "Red Bull" = "#0600ef",
                  "Ferrari" = "#dc0000",
                  "Racing Point" = "#F596C8",
                  "Force India" = "#f596c8",
                  "AlphaTauri" = "#ffffff",
                  "McLaren" = "#ff8700",
                  "Renault" = "#fff500",
                  "Williams" = "#0082fa",
                  "Toro Rosso" = "#469BFF",
                  "Lotus F1" = "#000000",
                  "Alfa Romeo" = "#960000",
                  "Sauber" = "#960000",
                  "Haas F1 Team" = "#787878")

key_teams <- c("Ferrari", 
               "McLaren",
               "Mercedes",
               "Red Bull",
               "Williams")

f1merged_hybrid %>%
  filter(positionText == 1) %>%
count(constructorname, sort = TRUE) %>% 
  ggplot(aes(x = n, 
             y = factor(constructorname, levels = rev(levels(factor(constructorname)))),
             fill = constructorname)) +
  geom_col(aes()) + 
 scale_fill_manual(values = team_colours) +
  labs(title = "Race Wins by Constructor",
       subtitle = "In the hybrid era (2014-2020)",
       x = "Race Wins",
      y = "Constructor") +
  guides(fill = "none")
```

Our bar chart shows Mercedes has an impressive amount of wins when compared to the other key teams. But why is this?

We thought it could possibly be due to Mercedes having more drivers leading to a misleading result. However, the merged dataset we are working with has the same amount of drivers for each key team. 

Could this be due to Mercedes being faster than the other constructors? We investigated the average lap time of each key team and found that Mercedes has a slightly lower median lap time than the other teams. However, this isn't a significant enough difference to explain their massive amount of wins - if it were, Ferrari and Red Bull would be close to Mercedes amount of wins.





Write-up of your project and findings go here. 
Think of this as the text of your presentation. 
The length should be roughly 5 minutes when read out loud.
Although pacing varies, a 5-minute speech is roughly 750 words. 
To use the word count addin, select the text you want to count the words of (probably this is the 
Summary section of this document, go to Addins, and select the `Word count` addin).
This addin counts words using two different algorithms, but the results should be similar and as long as you're in the ballpark of 750 words, you're good! 
The addin will ignore code chunks and only count the words in prose.

You can also load your data here and present any analysis results / plots, but I strongly urge you to keep that to a minimum (maybe only the most important graphic, if you have one you can choose). 
And make sure to hide your code with  `echo = FALSE` unless the point you are trying to make is about the code itself.
Your results with proper output and graphics go in your presentation, this space is for a brief summary of your project.

```{r load-data, echo = FALSE}
# load data here, if you like
```

## Presentation

Our presentation can be found [here](presentation/presentation.html).

## Data 

The data we have used in this project is from the Ergast Developer API, which has collected its data from the official race classifications released by the FIA (Formula One's governing body).

Ergast, A 2009, *Ergast Delevoper API*, data zip file,
The Ergast Developer API,
retrieved 22nd October 2021, <http://ergast.com/mrd/>

## References

Ergast, A 2009, *Ergast Delevoper API*, data zip file,
The Ergast Developer API,
retrieved 22nd October 2021, <http://ergast.com/mrd/>

