---
title: "What makes a successful Formula One team?"
author: by ECK
output: github_document
---

```{r, libraries-and-data, echo = FALSE, message = FALSE}
library(tidyverse)
library(broom)
library(here)
library(skimr)
library(glue)
library(forcats)

f1merged <- read_csv("/cloud/project/data/f1merged.csv")
f1merged_hybrid <- read_csv("/cloud/project/data/f1merged_hybrid.csv")
```


## Summary

Formula One, or F1, is the highest class of single-seater racing and one of the most prestigious racing series around the world. 

Each season around 10 constructors (also referred to as "teams") build a car to strict regulations and race them at more than 20 circuits across the globe. A constructor fields two drivers and earns points towards two separate championships; one each for the driver and constructor who score the most points over a season.

With our project we wanted to investigate the factors that make a successful F1 team. Is outright speed the only metric that matters? Or do small gains made in pit stop times or reliability make all the difference?

In picking out a successful Formula One team from recent years there is one obvious candidate.

```{r, wins-bar-chart, echo = FALSE}
team_colours <- c("Mercedes" = "#00d2be",
                  "Red Bull" = "#0600ef",
                  "Ferrari" = "#dc0000",
                  "Racing Point" = "#F596C8",
                  "Force India" = "#f596c8",
                  "AlphaTauri" = "#ffffff",
                  "McLaren" = "#ff8700",
                  "Renault" = "#fff500",
                  "Williams" = "#0082fa",
                  "Toro Rosso" = "#469BFF",
                  "Lotus F1" = "#000000",
                  "Alfa Romeo" = "#960000",
                  "Sauber" = "#960000",
                  "Haas F1 Team" = "#787878")

key_teams <- c("Ferrari", 
               "McLaren",
               "Mercedes",
               "Red Bull",
               "Williams")

f1merged_hybrid %>%
  filter(positionText == 1) %>%
count(constructorname, sort = TRUE) %>% 
  ggplot(aes(x = n, 
             y = factor(constructorname, levels = rev(levels(factor(constructorname)))),
             fill = constructorname)) +
  geom_col(aes()) + 
 scale_fill_manual(values = team_colours) +
  labs(title = "Race Wins by Constructor",
       subtitle = "In the hybrid era (2014-2020)",
       x = "Race Wins",
      y = "Constructor") +
  guides(fill = "none")
```

Mercedes-AMG Petronas F1 team have won the Drivers' and Constructors' championships 7 years running, with a near unprecedented level of dominance. We looked at Mercedes specifically, as well as a comparison group containing 4 other teams; Ferrari, McLaren, Red Bull and Williams. These comparison teams were chosen because they had all competed in F1 across the entire period we looked at, so we could easily compare them to Mercedes directly.

To answer the question "What makes a successful Formula One team?", we wanted to analyse how several predictor variables from our data set impacted a team's performance. Since our data is actually a combination of multiple data sets, we combined the data sets *results, races, drivers* and *constructors* and focused only on entries in the hybrid era (2014-2020 excluding the 2021 season which is ongoing at the time of this analysis). This resulting data set was called *f1merged_hybrid*.

Our bar chart shows Mercedes has an impressive amount of wins when compared to the other key teams. But why is this?

Could this simply be due to Mercedes being outright faster than their competitors? We investigated the average lap time of each key team and found that Mercedes has only a slight advantage in median lap time over the likes of Ferrari and Red Bull. Is this small advantage in race pace a significant enough difference to explain the massive discrepancy in wins and pole positions? (1st place in qualifying) Looking purely at median lap time, it wouldn't be unreasonable to expect Ferrari and Red Bull to have similarly successful records over the same period.

Do tiny margins in terms of speed make a significant difference in F1? And what factors other than outright speed might play into a winning season?



--------------------------------------------------------------------------------------------

Write-up of your project and findings go here. 
Think of this as the text of your presentation. 
The length should be roughly 5 minutes when read out loud.
Although pacing varies, a 5-minute speech is roughly 750 words. 
To use the word count addin, select the text you want to count the words of (probably this is the 
Summary section of this document, go to Addins, and select the `Word count` addin).
This addin counts words using two different algorithms, but the results should be similar and as long as you're in the ballpark of 750 words, you're good! 
The addin will ignore code chunks and only count the words in prose.

You can also load your data here and present any analysis results / plots, but I strongly urge you to keep that to a minimum (maybe only the most important graphic, if you have one you can choose). 
And make sure to hide your code with  `echo = FALSE` unless the point you are trying to make is about the code itself.
Your results with proper output and graphics go in your presentation, this space is for a brief summary of your project.

```{r load-data, echo = FALSE}
# load data here, if you like
```

## Presentation

Our presentation can be found [here](presentation/presentation.html).

## Data 

The data we have used in this project is from the Ergast Developer API, which has collected its data from the official race classifications released by the FIA (Formula One's governing body).

Ergast, A 2009, *Ergast Delevoper API*, data zip file,
The Ergast Developer API,
retrieved 22nd October 2021, <http://ergast.com/mrd/>

## References

Ergast, A 2009, *Ergast Delevoper API*, data zip file,
The Ergast Developer API,
retrieved 22nd October 2021, <http://ergast.com/mrd/>

