---
title: "Project proposal"
author: "ECK"
output: github_document
---

```{r load-packages, message = FALSE}
library(tidyverse)
library(broom)
```

## 1. Introduction
We will be looking at the question **"What makes a successful Formula One (F1) team?"**. In particular, we will look at Mercedes-AMG Petronas F1 Team, and their success in the hybrid era (2014-2020).

The data we are using comes from the Ergast Developer API (**http://ergast.com/mrd/**). It provides data for the Formula One racing series from 1950 to present. The data is collected from official race classifications released by the FIA, Formula One's governing body.

Much of our analysis will focus on the *results* dataframe. Each observation in this dataframe represents the result of one driver at one race. The variables are:  

+ Result ID   
+ Race ID   
+ Driver ID   
+ Constructor ID  
+ Driver number  
+ Starting grid position  
+ Official classification  
+ Points gained  
+ Laps completed  
+ Finishing time or gap  
+ Finishing time in milliseconds  
+ Lap number of fastest lap  
+ Fastest lap ranking relative to other drivers  
+ Fastest lap time  
+ Fastest lap speed   
+ Status ID  
 
 We will also look at several other data frames, both in conjunction with the *results* frame, and separately.
 
 These will include the *qualifying* data frame, with variables for race, driver, constructor & and position, as well as lap time.
 
Similarly we will use the *pit_stops* data frame, with variables for race, driver, stop number, lap number, time, duration, and duration in milliseconds.
 
We will also use the *drivers* and *constructors* dataframes to link driver names to their corresponding ID number.
 


## 2. Data

For this project, we will be using a combination of dataframes from the Formula One dataset. This combined dataframe will give us the variables referenced above in the **Introduction** section.

Here is a glimpse at one of the data frames we will be using called *results*.
```{r echo = FALSE}
results <- read_csv("/cloud/project/data/results.csv")
glimpse(results)
```


## 3. Data analysis plan

The plan for this data analysis is to explore the question   
**"What makes a successful Formula One (F1) team?"**

We will do this by wrangling and visualising our data to see if there is a relation between a Formula 1 team's performance and factors such as their pit stop data or their country of birth.
