---
title: "Project proposal"
author: "ECK"
output: github_document
---

```{r load-packages, message = FALSE}
library(tidyverse)
library(broom)
```

*For instructions on what each section should include, please see the [project page](https://idsed.digital/assessments/project/#proposal) on the course website. Remove this text when completing your proposal*.

## 1. Introduction
We will be looking at the question; What makes a successful Formula One (F1) team. In particular we will look at Mercedes-AMG Petronas F1 Team, and their success in the hybrid era (2014-2020).

The data we are using comes from the Ergast Developer API **http://ergast.com/mrd/**, it provides data for the Formula One racing series from 1950 to present. The data is collected from official race classifications released by the FIA, Formula One's governing body.

Much of our analysis will focus on the "results" dataframe. Each observation in this dataframe represents the result of one driver at one race. The variables are; 

Result ID, 
Race ID, 
Driver ID, 
Constructor ID,
Driver number,
Starting grid position,
Official classification,
Points gained,
Laps completed,
Finishing time or gap,
Finishing time in milliseconds,
Lap number of fastest lap,
Fastest lap ranking relative to other drivers,
Fastest lap time,
Fastest lap speed &
 Status ID
 
 We will also look at several other data frames, both in conjunction with the "results" frame, and separately.
 
 These will include the "qualifying" data frame, with variables for race, driver, constructor & position as well as lap time.
 
Similarly we will use the "pit_stops" data frame, with variables for race, driver, stop number, lap number, time, duration, and duration in milliseconds.
 
 


## 2. Data

For this project, we will be using a combination of dataframes from the Formula 1 dataset. This combined dataframe will give us the following:  
+ The drivers ID (*driverId*)  
+ The place they achieved in the race (*resultId*)  
+ **add more information here**

Here is a glimpse at the combined data frame we will be using. **(Change the dataframe here to the combined one once goal is decided)**
```{r echo = FALSE}
drivers <- read_csv("/cloud/project/data/drivers.csv")
glimpse(drivers)
```


## 3. Data analysis plan

The plan for this data analysis is to explore the question   
**"Question Here"**

We will do this by combining the dataframes **a, b, c...** and visualising **x, y, z**.

