---
title: "Project proposal"
author: "ECK"
output: github_document
---

```{r load-packages, message = FALSE}
library(tidyverse)
library(broom)
```

*For instructions on what each section should include, please see the [project page](https://idsed.digital/assessments/project/#proposal) on the course website. Remove this text when completing your proposal*.

## 1. Introduction

The dataframes we are using are from the website **website here** and has information on Formula 1 races. This includes variables such as the drivers information, the races they participated in, the race results and more.

## 2. Data

For this project, we will be using a combination of dataframes from the Formula 1 dataset. This combined dataframe will give us the following:  
+ The drivers ID (*driverId*)  
+ The place they achieved in the race (*resultId*)  
+ **add more information here**

Here is a glimpse at the combined data frame we will be using. **(Change the dataframe here to the combined one once goal is decided)**
```{r echo = FALSE}
drivers <- read_csv("/cloud/project/data/drivers.csv")
glimpse(drivers)
```


## 3. Data analysis plan

The plan for this data analysis is to explore the question   
**"Question Here"**

We will do this by combining the dataframes **a, b, c...** and visualising **x, y, z**.

