---
title: "Project proposal"
author: "ECK"
output: github_document
---

```{r load-packages, message = FALSE}
library(tidyverse)
library(broom)
library(here)
library(skimr)
```

## 1. Introduction
We will be looking at the question **"What makes a successful Formula One team?"**. In particular, we will look at Mercedes-AMG Petronas F1 Team, and their success in the hybrid era (2014-2020*). 

The data we are using comes from the Ergast Developer API (**http://ergast.com/mrd/**). It provides data for the Formula One (F1) racing series from 1950 to present. The data is collected from official race classifications released by the FIA, Formula One's governing body.

Much of our analysis will focus on the *f1merged* dataframe, which combines relevant data from the results, races, drivers & constructors data frames. Each observation in this dataframe represents the result of one driver at one race. 

Variables in this data frame include: 

+ Unique ID numbers and identifiers for each result, race, driver and constructor. 
+ The year/season, round number, date and name of each race.
+ Driver names and numbers. Constructor names and nationalities.
+ Starting and finishing positions for each driver.
+ Laps completed, points gained, finishing time.
+ Fastest lap time, fastest lap speed and fastest lap ranking.
+ Status ID, which links to a data frame with detailed finishing statuses.

We will also look at the *qualifying* data frame, with variables for race, driver, constructor and qualifying position, as well as fastest lap times from each qualifying session.
 
Similarly we will use the *pit_stops* data frame, with variables for race, driver, stop number, lap number, time of the pit stop and duration of the pit stop.
 
\**The 2021 season is excluded from our analysis because it is still ongoing at the time of writing*.

## 2. Data

For this project, we will be using a combination of data frames from the Formula One data set. The combined data frame *f1merged* will give us the variables referenced above in the **Introduction** section.

Here is a skim of the *f1merged* data frame.
```{r echo = FALSE}
f1merged <- read_csv("/cloud/project/data/f1merged.csv")
skim(f1merged)
```




## 3. Data analysis plan

The plan for this data analysis is to explore the question   
**"What makes a successful Formula One (F1) team?"**

To investigate this, we will look at the predictor variables('position' from the qualifying dataframe, 'duration', 'stop' from the pit_stops dataframe and 'fastestLapTime', 'fastestLapSpeed', 'rank', 'points', and 'positionText' from the results dataframe) and the outcome variable ('wins' from the constructor_standings dataframe) for Mercedes-AMG Petronas F1 Team during the hybrid era.

To do this we will compare the duration of a pit stop for the team throughout a season, and look at the efficiency and reliability of the pit crew. We will also compare the fastest lap time to other teams to explore the possible effect that updated engineering has had on the sport.

