---
title: "Project proposal"
author: "ECK"
output: github_document
---

```{r load-packages, message = FALSE}
library(tidyverse)
library(broom)
library(here)
library(skimr)
```

## 1. Introduction
We will be looking at the question **"What makes a successful Formula One (F1) team?"**. In particular, we will look at Mercedes-AMG Petronas F1 Team, and their success in the hybrid era (2014-2020).

The data we are using comes from the Ergast Developer API (**http://ergast.com/mrd/**). It provides data for the Formula One racing series from 1950 to present. The data is collected from official race classifications released by the FIA, Formula One's governing body.

Much of our analysis will focus on the *f1merged* dataframe, which combines the data we need from the results, races, drivers & constructors data frames. Each observation in this dataframe represents the result of one driver at one race. The variables are:  

+ Result ID   **(Add f1merged variables to this and expand on what they are)**
+ Race ID   
+ Driver ID   
+ Constructor ID  
+ Driver number  
+ Starting grid position  
+ Official classification  
+ Points gained  
+ Laps completed  
+ Finishing time or gap  
+ Finishing time in milliseconds  
+ Lap number of fastest lap  
+ Fastest lap ranking relative to other drivers  
+ Fastest lap time  
+ Fastest lap speed   
+ Status ID  
 
 We will also look at several other data frames, both in conjunction with the *f1merged* frame, and separately.
 
 These will include the *qualifying* data frame, with variables for race, driver, constructor & and position, as well as lap time.
 
Similarly we will use the *pit_stops* data frame, with variables for race, driver, stop number, lap number, time, duration, and duration in milliseconds.
 
 



## 2. Data

For this project, we will be using a combination of dataframes from the Formula One dataset. This combined dataframe will give us the variables referenced above in the **Introduction** section.

Here is a glimpse at one of the data frames we will be using called *f1merged*.
```{r echo = FALSE}
f1merged <- read_csv("/cloud/project/data/f1merged.csv")
glimpse(f1merged)
skim(f1merged)
```




## 3. Data analysis plan

The plan for this data analysis is to explore the question   
**"What makes a successful Formula One (F1) team?"**

To investigate this, we will look at the predictor variables('position' from the qualifying dataframe, 'duration', 'stop' from the pit_stops dataframe and 'fastestLapTime', 'fastestLapSpeed', 'rank', 'points', and 'positionText' from the results dataframe) and the outcome variable ('wins' from the constructor_standings dataframe) for Mercedes-AMG Petronas F1 Team during the hybrid era.

To do this we will compare the duration of a pit stop for the team throughout a season, and look at the efficiency and reliability of the pit crew. We will also compare the fastest lap time to other teams to explore the possible effect that updated engineering has had on the sport.

